\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric, fit, calc}
\usepackage{xcolor}
\usepackage{listings}

\newcommand{\sequent}[2]{#1 \vdash #2}
\newcommand{\rulebox}[3]{%
  \begin{array}{c}#1\\\hline #2\end{array}\;\text{\scriptsize #3}}


\begin{document}

\title{Naïve Backward Proof Search for First-Order Logic\\
       and a Closure-Lookahead Improvement}

\titlerunning{FOL Proof Search with Closure Lookahead}

\author{Ben Holland\inst{1}}
\authorrunning{B.\ Holland}
\institute{School of Information and Communication Technology,\\
           Griffith University, Gold Coast, Australia\\
           \email{ben.holland@griffithuni.edu.au}}

\maketitle

\begin{abstract}

First-order logic is only semi-decidable, so any sound and complete
proof procedure is allowed to run forever on some inputs. This report
implements Algorithm~2 from Hou's \textit{Fundamentals of Logic and
Computation}: a naïve backward proof search in sequent calculus LK
that schedules rules by a five-tier priority and guesses witness terms
for the quantifier rules. I add four improvements on top of the
baseline: duplicate-formula elimination, branch-local loop detection by
ancestor-sequent hashing, one-step closure look-ahead during rule
selection, and Herbrand-guided witness choice that prefers terms
already occurring on the opposite side of the sequent. Each was added
in response to a specific failure mode I observed in the baseline.
I evaluate both provers on 102 formulae across four datasets: classical
propositional tautologies from Mendelson and
Troelstra--Schwichtenberg, a subset of Pelletier's seventy-five
problems, formulae taken from the 3806ICT course materials, and a
hand-written family of stress cases. Under a uniform step budget the
improved prover solves 93 formulae against the baseline's 82, uses
about half as many rule applications overall, and runs three to four
times faster on the course and synthetic datasets. The propositional
fragment is unchanged in verdict and slightly slower in absolute time,
and the Pelletier set shows a per-step overhead from the look-ahead
once the candidate-rule count grows. The exercise illustrates the
trade-off between brute systematic search and lightweight heuristics,
and shows that semi-decidability still bites once the simple
inefficiencies have been removed.

\keywords{Automated theorem proving \and First-order logic \and Sequent calculus \and Backward proof search.}
\end{abstract}

\section{Introduction}
\label{sec:intro}


Automated reasoning is the algorithmic search for proofs and refutations
of formal statements, and it sits behind every model checker,
SMT solver, and proof assistant in current use. Propositional logic has
a decision procedure (SAT is NP-complete), but first-order logic with
quantifiers and function symbols is only semi-decidable: a sound and
complete prover has to be allowed to run forever on some inputs. The
usual response is a calculus that decomposes a goal into syntactically
simpler subgoals, applied either forwards (resolution, paramodulation)
or backwards (sequent calculus, tableaux). Gentzen's sequent calculus
LK~\cite{Gentzen1935} is a natural target for backward search because
its rules mirror the connectives of the language and a derivation can be
read directly off the search tree~\cite{Hou2021,TroelstraSchwichtenberg2000}.

Production theorem provers depart from the textbook calculus in almost
every direction: they convert to clausal form, apply ordered resolution
with paramodulation, prune redundant clauses, and steer the search with
hand-tuned heuristics. None of this is present in Algorithm~2 of
Hou~\cite[p.~67]{Hou2021}, which is intentionally minimal: a five-tier
rule priority, no contraction, no unification, no redundancy
elimination, and witness terms chosen by enumeration. The algorithm is
presented as a teaching artefact, and its performance on classical
benchmarks like Pelletier's~\cite{Pelletier1986} reflects that.

The question I take up in this report is how much of the gap between
Algorithm~2 and what a production prover does can be closed with small
self-contained changes that preserve the algorithm's overall shape.
I implement the textbook procedure in Python, identify four failure
modes by inspecting divergent search trees, and add four sound
improvements (I1--I4) that address them. I then evaluate baseline and
improved variants on 102 formulae from four sources: classical
propositional tautologies, the canonical Pelletier problems, the
course's quantifier exercises, and a hand-written family of stress
cases. Section~\ref{sec:approach} describes the calculus, the
baseline, the four improvements, and the hypotheses they should
satisfy. Section~\ref{sec:impl} covers the implementation, the
benchmark, and the experimental results. Section~\ref{sec:discussion}
reviews the remaining limitations and the obvious next steps.

\section{Proposed approach}
\label{sec:approach}


\subsection{Background: sequent calculus LK}
\label{sec:lk}


A sequent $\Gamma \vdash \Delta$ asserts that whenever every formula in
the antecedent multiset $\Gamma$ is true, at least one formula in the
succedent multiset $\Delta$ is true~\cite{TroelstraSchwichtenberg2000}.
Backward proof search starts from a goal sequent $\vdash\varphi$ and
applies LK rules in reverse, replacing each sequent with the premises
that would entail it; a branch closes when an axiom (\textsc{id},
\textsc{TR}, or \textsc{FL}) becomes applicable. The rules of LK fall into three groups (Hou~\cite[Ch.\,2]{Hou2021}). The
\emph{invertible non-branching} rules ($\wedge L$, $\vee R$, $\to R$,
$\neg L$, $\neg R$, $\forall R$, $\exists L$) decompose a single
connective without splitting and never lose information. The
\emph{invertible branching} rules ($\wedge R$, $\vee L$, $\to L$) split a
goal into two subgoals that must both close. The \emph{non-invertible}
quantifier rules ($\forall L$, $\exists R$) require a witness term and,
crucially, retain the principal formula in their premise; this is the
syntactic reason the search may not terminate, since a formula
$\forall x.\,A$ on the antecedent can be re-instantiated at every step.
Every rule preserves classical validity, and Gentzen's
\textit{Hauptsatz}~\cite{Gentzen1935} guarantees that a derivation
exists if and only if the goal is valid. The best one can hope for is
therefore a sound and complete procedure that may not terminate.


\subsection{Baseline: Algorithm 2}
\label{sec:baseline}


Algorithm~2 maintains a queue of open branches and, at each step,
applies the highest-priority rule that fires on the current sequent.
The priority schedule (Figure~\ref{fig:arch}) has five tiers: axioms
close a branch first; invertible non-branching rules decompose a
connective without splitting; invertible branching rules split a branch
in two; the non-invertible quantifier rules $\forall L$ and $\exists R$
are tried only when no invertible rule applies, first with terms
already used on the current branch and finally with a freshly
introduced constant. The order within a tier does not affect
provability because the invertible rules are confluent, so the priority
is an efficiency choice rather than a logical one. The algorithm is
sound by inspection, and complete in the sense that a derivation
exists for every classical theorem; in practice it is incomplete
because $\forall L$ and $\exists R$ keep their principal formula in the
premise, so the antecedent can grow without bound through repeated
instantiations.

A small example: on the goal
\[
(\forall x.\,P(x)\to Q(x))\to((\forall x.\,P(x))\to(\forall x.\,Q(x)))
\]
the procedure applies $\to R$ twice and then $\forall R$, reducing the
sequent to $\forall x.\,P(x)\to Q(x),\ \forall x.\,P(x) \vdash Q(c)$
for a fresh $c$. It then falls through to $\forall L$ with no
information about which witness to use, enumerates terms blindly, and
on three course-set theorems of this shape runs out of step budget
before hitting the right witness. Section~\ref{sec:improvements} targets exactly this
failure mode. To make the experiment finite I add two caps: a global
step limit of 300 rule applications and a per-formula cap of three
fresh constants per $\forall L$/$\exists R$ principal. Both are sound
under-approximations --- a formula reported as \emph{aborted} may
still be valid, but no valid formula is ever reported as
\emph{countermodel} --- and the values were picked so that the
first-order failures I see come from the algorithm, not from the
budget.

\begin{figure}[!htb]
  \centering
  \vspace{-2mm}
  \scalebox{0.78}{%
  \begin{tikzpicture}[
    node distance=4mm and 6mm,
    every node/.style={font=\small},
    box/.style={draw, rounded corners=2pt, align=center,
                 minimum width=34mm, minimum height=6mm},
    rule/.style={box, fill=blue!7},
    decision/.style={diamond, draw, aspect=2.2, align=center,
                     fill=yellow!10, minimum width=38mm, minimum height=5mm},
    arr/.style={-Latex},
    edgelabel/.style={font=\scriptsize, midway, fill=white,
                       inner xsep=2pt, inner ysep=1pt}
  ]
    \node[box, fill=gray!10] (in) {Input formula $\varphi$\\(parsed AST)};
    \node[box, below=of in] (init) {Initialise queue with root sequent\\
                                    $\,\vdash\varphi$};
    \node[decision, below=8mm of init] (q) {Open branch?};
    \node[rule, below=of q] (ax) {1.\ Axiom (id, TR, FL)};
    \node[rule, below=of ax] (nb) {2.\ Invertible non-branching\\
                                     $\wedge L,\ \vee R,\ \to R,\ \neg L,\ \neg R,\ \forall R,\ \exists L$};
    \node[rule, below=of nb] (br) {3.\ Invertible branching\\
                                     $\wedge R,\ \vee L,\ \to L$};
    \node[rule, below=of br] (qu) {4.\ Quantifier with unused term\\
                                     $\forall L,\ \exists R$};
    \node[rule, below=of qu] (qf) {5.\ Quantifier with fresh term\\
                                     (bounded by \textsc{fresh\_cap})};
    \node[box, fill=red!10, below=of qf] (stop) {6.\ Stop $\Rightarrow$ open leaf};
    \node[box, fill=green!10, right=24mm of q] (out) {All branches closed?\\
        \emph{proved} / \emph{countermodel} / \emph{aborted}};

    \draw[arr] (in) -- (init);
    \draw[arr] (init) -- (q);
    \draw[arr] (q) -- node[edgelabel,left]{yes} (ax);
    \draw[arr] (q.east) -- node[edgelabel,above]{no} (out.west);
    \draw[arr] (ax) -- (nb);
    \draw[arr] (nb) -- (br);
    \draw[arr] (br) -- (qu);
    \draw[arr] (qu) -- (qf);
    \draw[arr] (qf) -- (stop);
  \end{tikzpicture}}
  \caption{Baseline control flow (Hou 2021, Algorithm 2). Each iteration
  pulls an open branch from the queue and applies the highest-priority
  applicable rule. If no rule applies the branch closes as an open
  counter-model leaf. The \emph{Improved} variant retains the same
  five-level priority but instruments rules 2--4 with the four enhancements
  in Section~\ref{sec:improvements}.}
  \label{fig:arch}
\end{figure}

\subsection{Improvements over Algorithm 2}
\label{sec:improvements}


\paragraph{(I1) Duplicate-formula elimination (implicit contraction).}

Before each rule application I check whether the conclusion would
introduce a formula already present on the same side of the target
sequent and, if so, skip the application. This is the structural rule
of contraction applied implicitly: it is sound because LK admits
contraction as a derived rule, and it is helpful because the
non-invertible quantifier rules retain their principal formula and would
otherwise produce arbitrarily many copies of the same instance. The
optimisation costs an $O(|\Gamma|+|\Delta|)$ membership test per rule
application but bounds the size of any sequent above by the number of
distinct subformulae of the goal, removing one of the textbook
algorithm's main termination concerns.

\paragraph{(I2) Branch-local loop detection.}

Each open branch carries the set of canonical sequent signatures it has
already passed through, where a signature is the multiset of formulae
serialised under a fixed total order. Before applying a rule I hash the
candidate premise against this set, and a rule whose premise would
revisit an ancestor sequent is treated as inapplicable. The rule is
sound because LK admits weakening, so a sequent that closed once on its
ancestors closes again from below; in propositional logic the check
guarantees termination because every branch has a finite subformula
universe, and in first-order logic it cuts off the most common
quantifier loops produced by repeated $\forall L$ applications with the
same witness. Sibling branches do not share their visited sets, so
parallel cousins can still revisit each other; I accept this loss as
the price of the test being $O(1)$ on a hashed signature.

\paragraph{(I3) Closure-lookahead rule selection.}

Before committing to the highest-priority applicable rule I score every
applicable invertible rule by how many of its premises would close
immediately by an axiom on the current branch, and prefer rules whose
premises all close. Ties are broken by tier order, so a non-branching
rule wins over a branching rule with the same closure score. The check
is a one-step look-ahead and does not change provability, since the
chosen rule is already invertible and its premises would have been
generated eventually under the textbook schedule; the saving comes from
avoiding subgoals that the schedule would otherwise reach by a longer
path. The cost is a constant-factor overhead per decision, paid even on
trivial cases where any applicable rule would have closed the branch;
the experiment in Section~\ref{sec:results} exposes this trade-off on
the Pelletier dataset, where the candidate-rule count grows with the
predicate count and the overhead becomes visible.

\paragraph{(I4) Herbrand-guided quantifier instantiation.}

When instantiating $\forall L$ on a formula whose body has head
predicate $P$, the improved prover prefers terms already occurring as
arguments of $P$ on the opposite side of the sequent; the same
heuristic applies to $\exists R$ with the sides reversed. This is a
syntactic approximation of the Herbrand witness most likely to close
the resulting sequent by \textsc{id}: an \textsc{id} closure requires
the same atomic formula on both sides, so any such witness is
necessarily a Herbrand term already in play. The procedure is sound
provided the candidate term does not contain any variable bound
elsewhere in the principal formula; otherwise the substitution would
produce a renamed copy of the principal formula and the search would
loop. The bound-variable filter recurses into compound terms to
enforce this.

\subsection{Hypotheses}
\label{sec:hypotheses}


I commit to three predictions.
\textbf{H1 (no regression).} The set of formulae solved by the improved
prover is a superset of those solved by the baseline; soundness is
preserved by construction.
\textbf{H2 (gain concentrated on first-order).} The largest gains are
expected on the course and synthetic datasets, which contain
quantifier-shuffling theorems whose proofs require $\forall L$ to be
instantiated with a constant introduced by $\forall R$ on the same
branch --- exactly the case the Herbrand heuristic and closure
look-ahead were designed to address.
\textbf{H3 (overhead on the easy fragment).} On purely propositional
formulae the verdicts and step counts should be identical between the
two provers, but the closure look-ahead should add a small constant
overhead.

\section{Implementation and experiment}
\label{sec:impl}

\subsection{Implementation}
\label{sec:implementation}


The implementation is in Python 3.11+ using only the standard library;
no external solver or unification package is invoked. The source tree
separates concerns into five modules: the formula AST and capture-avoiding
substitution (\texttt{formula.py}), a recursive-descent parser for the
course syntax (\texttt{parser.py}), the baseline prover
(\texttt{baseline.py}), the improved prover (\texttt{improved.py}), and
the benchmark harness (\texttt{bench.py}). Sequents are represented as a
pair of Python lists with multiset semantics, and the priority schedule
is a single dispatch loop that short-circuits at the first applicable
tier. Capture-avoiding substitution follows the standard recursive
definition. Both provers expose the same
\texttt{prove(formula,\,step\_limit,\,fresh\_cap)} entry point.

\subsection{Datasets}
\label{sec:datasets}


\begin{itemize}
\item \textbf{Propositional} (25 formulae). Classical propositional
  tautologies and non-tautologies — Hilbert axiom schemas, de Morgan,
  Peirce's law, distributivity. Sources: Mendelson~\cite{Mendelson1997},
  Troelstra and Schwichtenberg~\cite{TroelstraSchwichtenberg2000}.
\item \textbf{Pelletier} (26 formulae). The propositional fragment
  (problems 1--17) and the first monadic / dyadic problems (18--25, 38),
  plus three known non-theorems. Source:
  Pelletier~\cite{Pelletier1986}.
\item \textbf{Course} (22 formulae). Workshop 3--5 exercises and Hou's
  Chapter 2 worked examples — quantifier de Morgan, distribution, swap.
  Source: Hou~\cite{Hou2021} and the 3806ICT workshop handouts.
\item \textbf{Synthetic} (29 formulae). Parameterised families designed to
  stress specific algorithmic behaviour: deep right-associated implication
  chains, deep $\wedge$/$\vee$ chains, quantifier-instantiation depth,
  redundant formulae (contraction tests), nested quantifier alternation,
  and Skolem-style function iteration. Hand-written for this assignment.
\end{itemize}


Each dataset targets a different aspect of the prover. The
propositional set isolates non-quantifier behaviour: every formula
closes in a few steps under either prover, so a divergence here would
flag an implementation bug rather than an algorithmic difference.
The Pelletier subset is a standard external benchmark for first-order
provers, which the marker can verify independently. The course set
covers the workshop and textbook material from 3806ICT directly.
The synthetic set is adversarial: each family is sized so that the
baseline's blind enumeration shows up clearly while the improved
prover's heuristics finish in single-digit steps. I labelled every
formula by hand against classical Tarski semantics, and each file
mixes formulae the baseline solves easily with formulae it does not.

\subsection{Experimental setup}
\label{sec:setup}


All measurements were taken on a Windows~11 laptop with an Intel Core
i7-1165G7 and 16\,GB of memory, running CPython~3.12.10. Each prover is
invoked once per formula with $\textsc{step\_limit}=300$ and
$\textsc{fresh\_cap}=3$, and wall-clock time is measured by
\texttt{time.perf\_counter()} around each \texttt{prove()} call. Step
counts are deterministic given a fixed \texttt{PYTHONHASHSEED}; the
reported figures are from a single run with \texttt{PYTHONHASHSEED=0}
and are reproducible by re-running \texttt{python -m src.bench}. The
analysis below relies primarily on step counts, which are bit-identical
across reruns under the same seed.

\subsection{Results}
\label{sec:results}

\begin{table}[!htb]
  \caption{Per-dataset comparison of Baseline vs Improved
  (I1--I4 layered) over 102 formulae; ``correct'' = matched ground-truth
  label, $\sum$steps and $\sum$ms are summed across the dataset.
  \textsc{step\_limit}=300, \textsc{fresh\_cap}=3.}
  \label{tab:results}
  \centering
  \begin{tabular}{lrrrrrrr}
    \toprule
    Dataset & $N$ & \multicolumn{2}{c}{Correct} & \multicolumn{2}{c}{$\sum$ steps} & \multicolumn{2}{c}{$\sum$ ms} \\
            &     & Base & Impr & Base  & Impr  & Base  & Impr  \\
    \midrule
    Propositional & 25 & 25 & 25 &  142 &  142 &    0.7 &    2.1 \\
    Pelletier     & 26 & 20 & 22 & 1688 & 1202 & 1651.5 & 7230.3 \\
    Course        & 22 & 14 & 18 & 2491 & 1324 & 1032.7 &  320.4 \\
    Synthetic     & 29 & 23 & 28 & 1910 &  509 & 1269.6 &  315.4 \\
    \midrule
    \textbf{Total} & \textbf{102} & \textbf{82} & \textbf{93}
                   & \textbf{6231} & \textbf{3177}
                   & \textbf{3954.6} & \textbf{7868.1} \\
    \bottomrule
  \end{tabular}
\end{table}


The improved prover solves 93/102 formulae against the baseline's 82,
with no regressions. Every divergence in Table~\ref{tab:divergences}
is a case the baseline aborts at the step limit and the improved
prover finishes in under seventy steps. Nine of the eleven are
quantifier-shuffling theorems whose proofs need $\forall L$
instantiated with a constant introduced earlier by $\forall R$ on the
same branch — the case the Herbrand heuristic targets directly.
The remaining two, \texttt{synthetic\#28} and \texttt{synthetic\#29},
involve iterated function symbols ($P(f^k(a))$ for $k\in\{2,3\}$) and
are dispatched by duplicate-formula elimination.

Total step count drops by 49\% (6231 to 3177), and on no dataset does
the improved prover use more steps than the baseline. Wall-clock time
is more mixed. The improved prover is 3.2$\times$ faster on the course
set and 4.0$\times$ faster on synthetic, dominated by the divergences
which previously took up to 767\,ms each. The propositional set picks
up a roughly 1.4\,ms penalty across 25 formulae from the look-ahead
overhead on cases where no rule selection is actually needed. Pelletier
is the only dataset where the improved prover is slower (7.2\,s versus
1.7\,s), even though it solves two more cases: problems 18--25 contain
many predicates, so the closure-look-ahead test runs over a larger
candidate-rule set on every decision and the per-decision overhead
accumulates.

The three hypotheses of Section~\ref{sec:hypotheses} hold up: H1
without exception, H2 on the course and synthetic datasets and only
partially on Pelletier, and H3 with the predicted small time penalty.
Overall the improvements add eleven solved cases and roughly halve the
search effort on the first-order benchmarks.

\begin{table}[!hbt]
  \caption{All 11 cases where the two provers disagree; in every case the baseline aborts at the step limit and the improved prover proves the formula in under 70 steps.}
  \label{tab:divergences}
  \centering
  \tiny
  \setlength{\tabcolsep}{3pt}
  \renewcommand{\arraystretch}{0.85}
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{lll rr}
    \toprule
    Case & Expected & Formula (abbrev.) & Base steps & Impr steps \\
    \midrule
    course\#4   & proved & $(\forall x.\,R_1(x){\to}R_2(x)){\to}((\forall x.\,R_1(x)){\to}(\forall x.\,R_2(x)))$ & 300 abort & 9 \\
    course\#12  & proved & $(\forall x\forall y.\,R(x,y)){\to}(\forall y\forall x.\,R(x,y))$       & 300 abort & 8 \\
    course\#13  & proved & $(\exists x\exists y.\,R(x,y)){\to}(\exists y\exists x.\,R(x,y))$       & 300 abort & 9 \\
    course\#18  & proved & $(\exists x\forall y.\,R(x,y)){\to}(\forall y\exists x.\,R(x,y))$       & 300 abort & 7 \\
    pelletier\#19 & proved & Pelletier 19 (existential conjunction)                                  & 300 abort & 41 \\
    pelletier\#22 & proved & Pelletier 22 (multi-quantifier with disjunctive premise)               & 300 abort & 67 \\
    synthetic\#19 & proved & $(\forall x.\,P(x){\to}Q(x)){\to}((\forall x.\,P(x)){\to}(\forall x.\,Q(x)))$ & 300 abort & 9 \\
    synthetic\#25 & proved & $(\exists x\forall y.\,R(x,y)){\to}(\forall y\exists x.\,R(x,y))$     & 300 abort & 7 \\
    synthetic\#26 & proved & $(\forall x\forall y.\,R(x,y)){\to}(\exists x\exists y.\,R(x,y))$     & 300 abort & 6 \\
    synthetic\#28 & proved & $(\forall x.\,P(x){\to}P(f(x)))\,\&\,P(a)\to P(f(f(a)))$              & 300 abort & 17 \\
    synthetic\#29 & proved & $(\forall x.\,P(x){\to}P(f(x)))\,\&\,P(a)\to P(f(f(f(a))))$           & 300 abort & 23 \\
    \bottomrule
  \end{tabular}}
\end{table}

\section{Discussion and conclusion}
\label{sec:discussion}


The improved prover hits the same semi-decidability ceiling as the
baseline: a formula reported as \emph{aborted} cannot be told apart
from a non-theorem. Loop detection is branch-local, so sibling
subgoals can revisit the same sequent independently of each other.
The Herbrand heuristic only looks at the immediate sequent and does
not pass witness suggestions across branches. And none of the four
improvements adds unification, so the prover cannot tell that two
sequents differ only in the choice of a metavariable.

The most useful next step would be to replace the cap-based termination
with iterative deepening on instantiation depth. Adding proper
unification — the Smullyan-style free-variable tableau treatment of
$\forall L$/$\exists R$ — would replace the Herbrand heuristic with
exact rather than approximate witness selection. Running the same
benchmark against a production prover would also show how much of the
remaining gap is unification and how much is everything else.

\paragraph{Data availability.}
Source code, datasets, and benchmark scripts are available at
\url{https://github.com/MysticalMrCool/3806ict-assignment1-2026}.

\paragraph{Generative AI acknowledgement.}
Anthropic Claude was used for coding help, LaTeX/TikZ formatting,
dataset-label checks, and rubric checks, not for writing report prose;
the prompts and responses are documented in Appendix~\ref{app:ai}.


\bibliographystyle{splncs04}
\bibliography{references}

\appendix

\section{Generative AI queries and responses}
\label{app:ai}


\input{appendix_ai}

\end{document}
